Interspecies comparisons of Mg/Ca ratios in limpet shells

### 2024-06-27 Manuscript_with_Heshkaia.qmd
first draft

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format: docx
#  elsevier-pdf:
#    keep-tex: true
#    journal:
#      name: Palaeo^3^
#      formatting: preprint
#      model: 3p
#      cite-style: authoryear
output:
  units: cm
---

@@

knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)
# Packages
  pacman::p_load(
    here,
    janitor,
    tidyverse,
    cowplot,ggx,ggpubr,
    patchwork,
    RColorBrewer)

@@

### Modern specimens

The analysed specimens, their origin and respective studies with research background information can be found in [Table 1](#Table_1). Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io]. While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.

::: {#Table_1}
| Context        | Study              | Species          | Location                                        | Sample ID     | Previous analyses                                           |
|------------|------------|------------|------------|------------|------------|
| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay - Tierra del Fuego (AR)          | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                                 | ND-1016-4     |                                                             |
|                |                    | *N. magellanica* |                                                 | NM-1016-1     |                                                             |
|                |                    |                  |                                                 | NM-1016-3     |                                                             |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - Westray (UK)                    | ORK-LT5       | Stable oxygen and carbon isotope analysis; Mg, Li, Sr, Ca,. |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - Westray (UK)                    | QG1-7188-1    | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                                 | QG1-7189-2    |                                                             |
|                |                    |                  |                                                 | QG2-1061-1    |                                                             |
|                |                    |                  |                                                 | QG2-1064-1    |                                                             |
|                |                    |                  |                                                 | QG2-7180-1    |                                                             |
|                |                    |                  |                                                 | QG2-7180-2    |                                                             |
|                |                    |                  |                                                 | QG1-7246-1    |                                                             |
|                | This study         | *N. magellanica* | Heshkaia 35 in Moat Bay - Tierra del Fuego (AR) | Heshkaia_35-1 | *none*                                                      |
|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-2 |                                                             |
|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-3 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-4 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-5 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-6 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-7 |                                                             |
|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-8 |                                                             |
|                |                    | *N. magellanica* |                                                 | Heshkaia_35-9 |                                                             |

@@

```{r SST_plot}
#| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
#| fig-height: 5
#| fig-width: 5
#| label: fig-SSTs

# Plot the data
SST_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "Monthly Sea Surface Temperature",
       x = "Month",
       y = "Mean Analysed SST (°C)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = c(0.1, 0.9))+  
  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) +
  labs(colour="",fill="")

@@

### Archaeological specimens

In total, we studied 16 archaeological specimens. Of those, seven *P. vulgata* come from Quoygrew on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba]. Together they show a range of lifespans, with some comparatively long for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to time-averaging. That said, the majority are under 5 years old. These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].

Nine additional shells of *N. magellanica* (n=6) and *N*. *deaureata* (n=3) have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp. shells. These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells. They derive from the site of Heshkaia 35 situated within Moat Bay, around 30 km east of where the modern specimens in our study were collected [@Zangrando2014-qi; @Zangrando2021-qg]. With radiocarbon dates putting the shells into the Late Holocene (1300 – 1450 CE) they are of similar age — if somewhat younger — as the assemblage from Quoygrew.

@@

```{r Nacella_Isotope_plot}
#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay (*a-f*).
#| fig-height: 10
#| fig-width: 10
#| label: fig-Nac_iso
# Filter data for Modern period
Nacella_iso <- D18O %>% filter(grepl("Nacella", species))

@@

Interestingly, the Mg/Ca intensity ratios in the *Nacella* sp. shells were much lower (0.05–0.30) than those seen in other patelloid species (e.g. 0.3–0.7 in *Patella vulgata* above, or 0.5–1.5 in ) using the same emission lines. This is particularly visible in @fig-Nac_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in [@Hausmann2023-ih]). While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere, or seen only in aragonitic parts of e.g. *Patella* shells.

@@

# Discussion

## Other correlations

```{r Ferguson Data}
#| fig-cap: Correlation graphs for Ferguson et al. specimens
#| fig-height: 20
#| fig-width: 10

Ferg_all <-
  Ferguson_data %>% 
  ggplot()+
  aes(d18o,mg_ca,col=shell)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))

@@

Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))

| Species       | Locality       | Specimen | Correlation R^2^        | Study                             |
|---------------|---------------|---------------|---------------|---------------|
| *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
|               |                | LAN545   | 0.86                    |                                   |
|               |                | LAN554   | 0.78                    |                                   |
|               |                | LAN559   | 0.82                    |                                   |
| *P. caerulea* | Croatia        | ISTPC1   | 0.9                     | [@Hausmann2019-fi]                |
|               |                | ISTPC2   | 0.84                    |                                   |
|               | Crete          | AF1911A  | 0.91[^1]                |                                   |
|               |                | AF3003A  | 0.92[^2]                |                                   |
|               | Israel         | AKKPC2   | 0.96                    |                                   |
|               |                | AKKPC3   | 0.89                    |                                   |
|               |                | FRMPC1   | 0.84                    |                                   |
|               |                | FRMPC2   | 0.96                    |                                   |
|               | Libya          | MO31A    | 0.83                    |                                   |
|               |                | MP64A    | 0.33                    |                                   |
|               |                | MP67A    | 0.96                    |                                   |
|               |                | MP68A    | 0.81                    |                                   |
|               | Malta          | MA10     | 0.82                    |                                   |
|               | Tunisia        | TUNPC1   | 0.81                    |                                   |
|               |                | TUNPC2   | 0.78                    |                                   |
|               | Turkey         | ANTPC1   | 0.95                    |                                   |
|               |                | ANTPC2   | 0.93                    |                                   |
|               |                | KIZPC1   | 0.94                    |                                   |
|               |                | KIZPC2   | 0.86                    |                                   |
| *P. rustica*  | Gibraltar      | JL1      | 0.02                    | [@Ferguson2011-zl]                |
|               |                | JL2      | 0.8 (0.79)              |                                   |
| *P. caerulea* | Gibraltar      | JM00     | 0.69 (0.79)             |                                   |
|               |                | JM30     | 0.83 (0.79)             |                                   |
| *P. vulgata*  | Orkney         | ORK-LT5  | not reported, here 0.88 | [@Graniero2017-io] and this study |
|               |                |          |                         |                                   |
|               |                |          |                         |                                   |
|               |                |          |                         |                                   |

: Table 2: Overview of comparative correlations {#tab:correlations}

[^1]: SST only, no other geochemical data available

[^2]: SST only, no geochemical data available

## Comparison of ORK-LT5

```{r}

P1 <-
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  # filter(std<0.1) %>% 
  # filter(mg_ca<0.7) %>% 
  # filter(dist<20) %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path(col="firebrick")+
  # xlim(0,10)+
  facet_wrap(~sampleID,scales = "free",nrow=1)

P2 <-
  D18O %>% 
  filter(sampleID=="ORK_LT5") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path(col="cornflowerblue")+
  scale_y_reverse()+
  facet_wrap(~sampleID,scales = "free",nrow=1)

plot_compare <- P1/P2

plot_compare

```{r subsample}

@@

@@

P3.1 <-
  MgCa_subsampled_150 %>% 
ggplot()+
  aes(mean_dist,mean_mg_ca)+
  geom_path(col="firebrick2")

P3.2 <-
  MgCa_subsampled_300 %>% 
ggplot()+
  aes(mean_dist,mean_mg_ca)+
  geom_path(col="firebrick4")

P1/P3.1/P3.2

@@

```{r}
# This is data extracted from Fig. 6E of Graniero et al. 2016
GranieroDF <- tibble::tribble(
                ~dist, ~mg_ca,
                  0.1,     19,
                  0.2,   20.6,
                  0.4,   23.8,
                  0.6,   20.4,
                    1,   19.1,
                  1.3,   14.9,
                  1.6,     16,
                  1.9,   17.8,
                  2.3,   15.8,
                  2.8,   17.3,
                  3.1,   13.8,
                  3.3,   24.9,
                  3.7,   16.1,
                    4,   15.6,
                  4.3,   18.4,
                  4.6,   15.5,
                  5.2,   18.6,
                  5.2,   21.2,
                  5.5,   24.1,
                  5.8,   18.7,
                  6.2,   19.3,
                  6.3,   15.5,
                  6.7,   16.7,
                  6.9,   15.3,
                  7.3,   17.9,
                  7.6,   13.7,
                    8,   11.8,
                  8.2,   16.2,
                  8.5,   18.1,
                  8.9,     18,
                  9.1,   19.7,
                  9.4,   20.8,
                  9.7,   19.6,
                   10,   22.2,
                 10.2,   20.2,
                 10.5,   19.3,
                 10.8,   22.8,
                 11.2,   18.7,
                 11.8,   22.1
                )

<!--# Maybe get the actual data from the image itself. -->

@@

```{r Modern iso ranges}
# Ranges <- D18O %>%
#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
# group_by(sampleID,period) %>%
#   mutate(max=max(d18O),
#          min=min(d18O)) %>%
#   select(min, sampleID,max) %>% unique()
# 
# Modern_Ranges <- Ranges %>%
#   filter(period=="Modern") %>%
#   arrange(range)

# Interestingly, the overall annual range in δ^18^O-values on Orkney (\~`r Modern_Ranges$range[3]`‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp. shells (`r min(Modern_Ranges$range)`‰ to `r max(Modern_Ranges$range)`‰) despite the different SST ranges.

```{r Correlation Graphs, eval=FALSE }
#| fig-cap: Correlation graphs for all specimens
#| fig-height: 20
#| fig-width: 10

P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  mutate(region = substr(sampleID, 1, 1)) %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
  # facet_wrap(~region,ncol=2,scales = "free")

>>>>

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of modern and archaeological samples (n=12) from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
---

@@

knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)
# Packages
  pacman::p_load(
    here,
    janitor,
    tidyverse,
    cowplot,ggx,ggpubr,
    patchwork,kableExtra,
    RColorBrewer)

@@

The analysed specimens, their origin and respective studies with research background information can be found in Table[1](#Table_1). Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io]. While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.

@@

::: {#Table_1}
| Context        | Study              | Species          | Location                               | Sample ID  | Previous analyses                                           |
|------------|------------|------------|------------|------------|------------|
| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay - Tierra del Fuego (AR) | ND-1016-3  | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                        | ND-1016-4  |                                                             |
|                |                    | *N. magellanica* |                                        | NM-1016-1  |                                                             |
|                |                    |                  |                                        | NM-1016-3  |                                                             |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - Westray (UK)           | ORK-LT5    | Stable oxygen and carbon isotope analysis; Mg, Li, Sr, Ca,. |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - Westray (UK)           | QG1-7188-1 | Stable oxygen and carbon isotope analysis                   |
|                |                    |                  |                                        | QG1-7189-2 |                                                             |
|                |                    |                  |                                        | QG2-1061-1 |                                                             |
|                |                    |                  |                                        | QG2-1064-1 |                                                             |
|                |                    |                  |                                        | QG2-7180-1 |                                                             |
|                |                    |                  |                                        | QG2-7180-2 |                                                             |
|                |                    |                  |                                        | QG1-7246-1 |                                                             |

```{r SST_plot,fig.pos='H'}
#| fig-cap: Monthly Sea Surface Temperature values averaged from  years. SST values were sourced from @Good2020-nl .
#| fig-height: 3
#| fig-width: 5
#| label: fig-SSTs

# Plot the data
SST_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "",
       x = "Month",
       y = "Mean Analysed SST (°C)",colour="",fill="") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = c(0.1, 0.9))+  
  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15))

@@

In total, we studied 7 archaeological specimens of *P. vulgata* coming from Quoygrew on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba]. Together they show a range of lifespans, with some comparatively long for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to time-averaging. That said, the majority are under 5 years old. These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].

@@

```{r Nacella_Isotope_plot}
#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay (*a-f*).
#| fig-height: 7
#| fig-width: 8
#| label: fig-Nac_iso
# Filter data for Modern period
Nacella_iso <- D18O %>% filter(grepl("Nacella", species))

@@

Interestingly, the Mg/Ca intensity ratios in the *Nacella* sp. shells were much lower (0.05–0.30) than those seen in other patelloid species (e.g. 0.3–0.7 in *Patella vulgata* above, or 0.5–1.5 in ) using the same emission lines. This is particularly visible in @fig-Nac_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in [@Hausmann2023-ih]). While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere, or seen only in aragonitic parts of e.g. *Patella* shells. Similar observations have been made before [@Graniero2017-io].

@@

# Discussion {#Discussion}

The Mg/Ca ratios from the 2D elemental imaging present positive results on SST being the main factor behind Mg/Ca concentration in three Atlantic limpet species. Annually repeating patterns of increasing (spring/summer) and decreasing (autum/winter) SST were also visible in the line scans, albeit less straightforward than in the 2D-data. Nevertheless, the resulting R^2^-values for each species were high (*P. vulgata mean R^2^*=0.89, *N. deaureata mean R^2^*=0.86*, N. magellanica mean R^2^*=0.88). These results are similar to previously published data on *Patella* sp. shells (Table [2](#Table_2)). R^2^ values of *P. depressa* lie between 0.78 and 0.87 [@Garcia-Escarzaga2021-ij], for *P. caerulea* from the Mediterranean between 0.78 and 0.96 — although with one outlier of 0.33, which had increased Mg/Ca ratios in increments with rich organic contents [@Hausmann2019-fi].

*P. rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.8, respectively [@Ferguson2011-zl] (@fig-Ferg). Together with the data from two *P. caerulea* shells (JM00 and JM30 with R^2^-values of 0.69 and 0.83), the data from JL2 was used determine a general Mg/Ca-SST relationship with an R^2^-value of 0.79. JL1 was disregarded in that calculation, as well as the second year of growth in JM00, which produced anomalous data. While no additional information about those specimens is available, it seems likely that the Mg/Ca ratios from JM00, which were interpreted as an ontogenetic trend, were potentially influenced by the same intra-increment heterogeneity we see in some of our shell specimens as well. Since these shells were analysed by homogenising the carbonate powder of the m+2 layer and the more exterior m+3 layer, these heterogeneities could potentially increased the Mg/Ca ratio unpredictably, leading to what might look like an ontogenetic trend in a linear dataset. Whether these anomalies affected JL1 in the same way, albeit stronger, is not clear but it might well be possible.

::: {#Table_2}
| Species       | Locality       | Specimen | R^2^                    | Study                             |
|--------------|--------------|--------------|--------------|-----------------|
| *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
|               |                | LAN545   | 0.86                    |                                   |
|               |                | LAN554   | 0.78                    |                                   |
|               |                | LAN559   | 0.82                    |                                   |
| *P. caerulea* | Croatia        | ISTPC1   | 0.9                     | [@Hausmann2019-fi]                |
|               |                | ISTPC2   | 0.84                    |                                   |
|               | Crete          | AF1911A  | 0.91                    |                                   |
|               |                | AF3003A  | 0.92                    |                                   |
|               | Israel         | AKKPC2   | 0.96                    |                                   |
|               |                | AKKPC3   | 0.89                    |                                   |
|               |                | FRMPC1   | 0.84                    |                                   |
|               |                | FRMPC2   | 0.96                    |                                   |
|               | Libya          | MO31A    | 0.83                    |                                   |
|               |                | MP64A    | 0.33                    |                                   |
|               |                | MP67A    | 0.96                    |                                   |
|               |                | MP68A    | 0.81                    |                                   |
|               | Malta          | MA10     | 0.82                    |                                   |
|               | Tunisia        | TUNPC1   | 0.81                    |                                   |
|               |                | TUNPC2   | 0.78                    |                                   |
|               | Turkey         | ANTPC1   | 0.95                    |                                   |
|               |                | ANTPC2   | 0.93                    |                                   |
|               |                | KIZPC1   | 0.94                    |                                   |
|               |                | KIZPC2   | 0.86                    |                                   |
| *P. rustica*  | Gibraltar      | JL1      | 0.02                    | [@Ferguson2011-zl]                |
|               |                | JL2      | 0.8\*\*                 |                                   |
| *P. caerulea* | Gibraltar      | JM00     | 0.69\*\*                |                                   |
|               |                | JM30     | 0.83\*\*                |                                   |
| *P. vulgata*  | Orkney         | ORK-LT5  | not reported, here 0.88 | [@Graniero2017-io] and this study |

: Table showing correlations of Mg/Ca and δ^18^O found in other studies. \* indicates that only SST data and no other geochemical data was used. \*\* these samples were used to determine a shared R^2^-value of 0.79.
:::

```{r Ferguson Data}
#| fig-cap: Correlation graphs for Ferguson et al., 2011 specimens.
#| fig-height: 20
#| fig-width: 10
#| label: fig-Ferg

Ferg_all <-
  Ferguson_data %>% 
  ggplot()+
  aes(d18o,mg_ca,col=shell)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca (mmol/mol)")+
  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))

@@

Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
```

Another shell whose previous results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with δ^18^O or SST in implication [@Graniero2017-io]. Different to other specimens of this study, we were able to resample this specimen and getting Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub). These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August). Including these missing parts lead to a better alignment of both records and a resulting R^2^-value of 0.88.

```{r ORK_LT5_Subsample}
#| fig-cap: Comparison of previous and new Mg/Ca data for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca and δ^18^O-values
#| fig-height: 10
#| fig-width: 15
#| label: fig-ORK_sub

@@

#
# This is data extracted from Fig. 6E of Graniero et al. 2016
GranieroDF <- tibble::tribble(
                ~dist, ~mg_ca,
                  0.1,     19,
                  0.2,   20.6,
                  0.4,   23.8,
                  0.6,   20.4,
                    1,   19.1,
                  1.3,   14.9,
                  1.6,     16,
                  1.9,   17.8,
                  2.3,   15.8,
                  2.8,   17.3,
                  3.1,   13.8,
                  3.3,   24.9,
                  3.7,   16.1,
                    4,   15.6,
                  4.3,   18.4,
                  4.6,   15.5,
                  5.2,   18.6,
                  5.2,   21.2,
                  5.5,   24.1,
                  5.8,   18.7,
                  6.2,   19.3,
                  6.3,   15.5,
                  6.7,   16.7,
                  6.9,   15.3,
                  7.3,   17.9,
                  7.6,   13.7,
                    8,   11.8,
                  8.2,   16.2,
                  8.5,   18.1,
                  8.9,     18,
                  9.1,   19.7,
                  9.4,   20.8,
                  9.7,   19.6,
                   10,   22.2,
                 10.2,   20.2,
                 10.5,   19.3,
                 10.8,   22.8,
                 11.2,   18.7,
                 11.8,   22.1
                )

@@

P3.1 <-  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca)+
  geom_path(show.legend = FALSE)+
    scale_x_continuous(breaks = seq(0,12,by=2),
                     labels = seq(0,12,by=2))+
  theme_cowplot(20)+
    annotate("rect", xmin = 1.5, xmax = 4, ymin = 0.3, ymax = 0.55, alpha = 0.2, fill = "grey50") +
  annotate("rect", xmin = 0, xmax = 0.5, ymin = 0.3, ymax = 0.55, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")

# P3.2 <-
#   MgCa_subsampled_300 %>% 
# ggplot()+
#   aes(mean_dist,mean_mg_ca)+
#   geom_path(col="firebrick4")

P3.3 <-
  GranieroDF %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path(col="black")+
   theme_cowplot(20)+
  scale_x_continuous(breaks = seq(0,12,by=2),
                     labels = seq(0,12,by=2))+
  annotate("rect", xmin = 1.5, xmax = 3, ymin = 10, ymax = 25, alpha = 0.2, fill = "grey50") +
  annotate("rect", xmin = 0, xmax = 0.08, ymin = 10, ymax = 25, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
  labs(x = "Distance to edge [mm]", y = "Mg/Ca (mmol/mol)")

P3.1/P3.3

@@

While we were not able to resample the *Nacella* sp. specimens in Graniero *et al.* [-@Graniero2017-io], we were able to source other modern and archaeological specimens of the same species from Tierra del Fuego, which can serve as an indicator. Similar to other patelloid shells, the results of the *Nacella* specimens in our study were successful with high R^2^-values (0.76--0.95) and suggest Mg/Ca as a reliable proxy for SST (@fig-Nac_Comp and supplementaries). Why these results are different from the studied *Nacella* sp. specimens in Graniero et al. [-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and δ^18^O records of their shells (Figure 6 in [@Graniero2017-io] ) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — fell apart and lead to an overall result of an uncorrelated relationship between Mg/Ca and SST.

# Conclusion

This study presents high-resolution and 2D information on Mg/Ca concentrations in modern and archaeological patelloid shells and their relation to sea surface temperature. The results for *Patella vulgata*, *Nacella deaureata*, and *Nacella magellanica* were consistently promising with mean R^2^-values of 0.89, 0.86, and 0.88 respectively.

LIBS elemental imaging provided some insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the δ^18^O of the same shells. Our 2D approach allowed us to avoid these areas of intra-increment heterogeneity and focus on areas that are predominantly controlled by changes in SST.

These results are particularly promising for future analyses of patelloid shells for the study of season of collection in archaeological sites as well as the reconstruction of sub-annual climatic conditions in the past. Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a large part of marine mollusc remains in archaeological sites of those regions. Improving their analysis in terms of cost and time efficiency through the benefits of LIBS, which can be used in this context as a screening tool for targeted δ^18^O analysis focusing on annual minima and maxima revealed in the Mg/Ca record, allows researchers to increase sample numbers and produce larger and more in-depth climatic studies than previously possible.

@@
